Using spatially biased variables in ecosystem condition accounting with a GIS based workflow

### 2024-01-28 manuscript/HIAs.qmd
clean script and add annotations

---
title: Using spatially biased variables in ecosystem condition accounting with a GIS based workflow
subtitle: A Short Subtitle
author:
  - name: Anders Lorentzen Kolstad
    email: anders.kolstad@nina.no
    affiliations: 
        - id: some-tech
          name: Norwegian Institute for Nature Research
          department: Department of Terrestrial Ecology
          address: Street Address
          city: City
          state: State
          postal-code: Postal Code
    attributes:
        corresponding: true
    note: This is the first author footnote.
  - name: Bob Security
    email: bob@example.com
    affiliations:
        - id: another-u
          name: Another University
          department: Department Name
          address: Street Address
          city: City
          state: State
          postal-code: Postal Code
    note: |
      Another author footnote, this is a very long footnote and it should be a really long footnote. But this footnote is not yet sufficiently long enough to make two lines of footnote text.
  - name: Cat Memes
    email: cat@example.com
    affiliations:
        - ref: another-u
    note: Yet another author footnote.
  - name: Derek Zoolander
    email: derek@example.com
    affilations:
        - ref: some-tech
abstract: |
  This is the abstract. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Vestibulum augue turpis, dictum non malesuada a, volutpat eget velit. Nam placerat turpis purus, eu tristique ex tincidunt et. Mauris sed augue eget turpis ultrices tincidunt. Sed et mi in leo porta egestas. Aliquam non laoreet velit. Nunc quis ex vitae eros aliquet auctor nec ac libero. Duis laoreet sapien eu mi luctus, in bibendum leo molestie. Sed hendrerit diam diam, ac dapibus nisl volutpat vitae. Aliquam bibendum varius libero, eu efficitur justo rutrum at. Sed at tempus elit.
keywords: 
  - SEEA EA
  - alien species
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: Ecological Indicators
      formatting: preprint
      model: 3p
      cite-style: authoryear
    include-in-header:
      text: '\usepackage{lineno}\linenumbers'
editor: 
  markdown: 
    wrap: sentence
# remove this execute part to actually run the analysis
execute:
  echo: false
  include: false
  eval: false
  warning: false
  message: false
---

```{r setup, warning=F, message=F, include=F}
#| warning: false
#| message: false
#| include: false
#| eval: true

library(tidyverse)
library(knitr)
library(sf)
library(tmap)
library(stars)
library(terra)
library(tidyterra)

myCRS <- 25832

@@

```{r paths}
# Conditional file directory
dir <- substr(getwd(), 1,2)

mireFile <- "P:/41201785_okologisk_tilstand_2022_2023/data/Myrmodell/myrmodell90pros.tif"

path_infrastructure <- ifelse(dir == "C:", 
        "R:/GeoSpatialData/Utility_governmentalServices/Norway_Infrastructure_Index/Original/Infrastrukturindeks_UTM33/infra_tiff.tif",
        "/data/R/GeoSpatialData/Utility_governmentalServices/Norway_Infrastructure_Index/Original/Infrastrukturindeks_UTM33/infra_tiff.tif")
```

@@

# Set up conditional file paths
# The data is too big to be stored on GitHub
dir <- substr(getwd(), 1,2)

path <- ifelse(dir == "C:", 
               "R:/GeoSpatialData/Habitats_biotopes/Norway_Miljodirektoratet_Naturtyper_nin/Original/Naturtyper_nin_0000_norge_25833_FILEGDB/Naturtyper_nin_0000_norge_25833_FILEGDB.gdb",
               "/data/R/GeoSpatialData/Habitats_biotopes/Norway_Miljodirektoratet_Naturtyper_nin/Original/Naturtyper_nin_0000_norge_25833_FILEGDB/Naturtyper_nin_0000_norge_25833_FILEGDB.gdb")

#pData <- ifelse(dir == "C:", 
#               "P:/41201785_okologisk_tilstand_2022_2023/data/",
#               "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/")

# Impart polygon data set
naturetypes <- sf::st_read(dsn = path)
# 117k polygons (2023)
# sometimes this functin struggles with Norwegian letters. Restart R and try again.

@@

# 13k obs.
#naturetypes |>
#  saveRDS("../data/surveyData.RDS")
#head(naturetypes)
```

@@

path <- paste0("../data/Basisdata_0000_Norge_25833_Kommuner_FGDB.gdb")

# find the correct layer
#st_layers(path)
muni <- sf::read_sf(path, layer = "kommune") |>
  st_transform(myCRS)

# This returns some multisurfaces as well
#table(st_geometry_type(muni))

# function to make sure that multipolygons are returned 
ensure_multipolygons <- function(X) {
    tmp1 <- tempfile(fileext = ".gpkg")
    tmp2 <- tempfile(fileext = ".gpkg")
    st_write(X, tmp1)
    gdalUtilities::ogr2ogr(tmp1, tmp2, f = "GPKG", nlt = "MULTIPOLYGON")
    Y <- st_read(tmp2)
    st_sf(st_drop_geometry(X), geom = st_geometry(Y))
}

@@

# To crop EDM, I need the three municipalities seprately. 
nf <- muni3 |>
  filter(kommunenummer == "3020")
tr <- muni3 |>
  filter(kommunenummer == "5001")
me <- muni3 |>
  filter(kommunenummer == "5418")

# I need to intersect nature types with municipality
nature3 <- naturetypes |>
  st_intersection(muni3)

@@

# and also to get the data coverage polygon
dk <-   read_sf("../data/dekningskartNF/Naturtyper NiN - Dekningskart.shp") |>
  rbind(read_sf("../data/dekningskartTR/Naturtyper NiN - Dekningskart.shp")) |>
  rbind(read_sf("../data/dekningskartME/Naturtyper NiN - Dekningskart.shp")) |> 
  st_transform(st_crs(muni3)) |>
  st_intersection(muni3) |>
  st_union()

@@

# calculate stats
# terrestrial area
terrestrial <- terrestrial |>
  mutate(area_t = geometry |> st_area())

```{r makeMap}
#| eval: false

tmap_mode("plot")
# Make map to show where the three municipalities are
# First, get centroids
centroids <- muni3 |>
  st_centroid()

positionMap <- tm_shape(outline)+
  tm_polygons() +
  tm_shape(centroids)+
    tm_text(text = "Municipality",
      size = 1,
      xmod = 2,
      ymod=-1)+
  tm_symbols(size = 2,
             shape = 17,
             col="black")+
  tm_grid(projection = 4326)+
  tm_layout(bg.color = "skyblue")

# Plot (could perhaps add the EDM here as well)
muniPlot <- tm_shape(muni3)+
    tm_borders()+
  tm_facets(by = "Municipality",
            ncol=2)+
  tm_shape(terrestrial)+
    tm_fill(col = "lightgreen",
             alpha=.4)+
  tm_shape(ocean)+
    tm_polygons(col = "skyblue",
                border.col = "black")+
  tm_shape(dk)+
    tm_polygons(col = "grey",
                alpha=.8)+
  tm_shape(nature3)+
  tm_polygons(col = "red",
              border.col = "red",
              lwd=3)

methodsMap <- tmap_arrange(positionMap, muniPlot,
             widths = c(.4, .6))

@@

# TEsting the cropping proceedure using a downsampled starts object 
mire_test <- read_stars(mireFile, proxy = T) |>
  setNames("mire") |>
  st_downsample(100) |>
  mutate(mire = na_if(mire, 0)) |>
  st_as_stars()

@@

# The cropping is correct:
#tmap_mode("plot")
#tm_shape(mire_test)+
#  tm_raster(palette = "red")+
#  tm_shape(outline)+
#  tm_borders()+
#  tm_shape(mire_test_nf)+
#  tm_raster(col="red")
```

```{r mireNF-error}
#| eval: false
# Then we do the same, but without downsampling

mire_nf <- read_stars(mireFile, proxy = T) |>
  setNames("mire") |>
  #mutate(mire = na_if(mire, 0)) |>
  st_crop(nf) |>
  st_as_stars()

# The cropped raster is for some reason far into the ocean. 
# The cells have values, so it seems like there is something 
# that happens after the cropping
tmap_mode("plot")
tm_shape(mire_test)+
  tm_raster(palette = "red")+
  tm_shape(outline)+
  tm_borders()+
  tm_shape(mire_nf)+
  tm_raster(col="red")

st_crs(mire_nf) <- 25832
```

```{r mireTerra}
# trying terra
# convert municipal outline to vect via st
nf_vect <- as(nf, "Spatial") |>
  terra::vect()

@@

```{r dk2}
# calculate terrestrial area
terrestrial <- terrestrial |> 
  mutate(t_area_km = 
           round(units::drop_units(area_t * 1e-6))) |>
  as_tibble()

# calculate area of survey coverage maps
dk2 <-   read_sf("../data/dekningskartNF/Naturtyper NiN - Dekningskart.shp") |>
  rbind(read_sf("../data/dekningskartTR/Naturtyper NiN - Dekningskart.shp")) |>
  rbind(read_sf("../data/dekningskartME/Naturtyper NiN - Dekningskart.shp")) |>
  st_transform(st_crs(muni3)) |>
  st_intersection(muni3) |>
  st_cast("POLYGON") |>
  group_by(Municipality) |>
  summarise(geometry = st_union(geometry)) |>
  mutate(dk_area_km = geometry |> st_area(),
         dk_area_km = round(units::drop_units(dk_area_km * 1e-6)))
```

```{r mireArea}
mireArea <- mire_nf_terra |>
  global(c("mean", "sum"), na.rm=T) |>
  add_column("Municipality" = "Nordre Follo") |>
  mutate(mirePercent = round(mean*100, 1),
         sum = sum/1e+4) |>
  dplyr::rename(mire_km2 = sum)

mire_nf_terra_dk <- mire_nf_terra |>
  terra::mask(dk2 |> filter(Municipality == "Nordre Follo")) |>
  global("sum", na.rm=T) |>
  mutate(mireInSurvey_km2 = sum/1e+4) |>
  add_column(Municipality = "Nordre Follo")

@@

```{r infrastructureIndex}
# this data is on a 100x100m grid
infra <- stars::read_stars(path_infrastructure) |>
  # , which is mor then we need - warp it to 1x1km
  st_warp(cellsize = c(1000, 1000), 
                                   crs = st_crs(nf), 
                                   use_gdal = TRUE,
                                   method = "average") |>
  setNames("infrastructureIndex") |>
  st_transform(myCRS) |>
   mutate(infrastructureIndex = case_when(
    infrastructureIndex < 1 ~ 0,
    infrastructureIndex < 6 ~ 1,
    infrastructureIndex < 12 ~ 2,
    infrastructureIndex >= 12 ~ 3)) |>
  st_crop(outline)

@@

path <- ifelse(dir == "C:", 
        "P:/41201785_okologisk_tilstand_2022_2023/data/cache/",
        "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/")

saveRDS(infra, paste0(path, "infrastructureIndex_discrete_vectorized.rds"))
```

```{r muni-tbl}
# number and area of mire polygons
nature3_tbl <- nature3 |>
  group_by(Municipality) |>
  summarise(n = n()) |>
  as_tibble()

muni_tbl <- muni3 |>
  # calculate total area
  mutate(area_km = 
           round(
             units::drop_units(
               geom |> st_area() * 1e-6))) |>
  # make tibble for the coming join
  as_tibble() |>
  # paste inn terrestrial area
  left_join(terrestrial |> select(kommunenummer, t_area_km),
            keep=F) |>
  # area of survey
  left_join(dk2 |> select(Municipality, dk_area_km)) |>
  mutate(dk_percent = round((dk_area_km/t_area_km) * 100)) |>
  # number of polygons
  left_join(nature3_tbl |> select(Municipality, n)) |>
  # total mire area and %
  left_join(mireArea |> select(Municipality, mirePercent, mire_km2)) |>
  # % mire inside survey coverage map
  left_join(mire_nf_terra_dk |> select(Municipality, mireInSurvey_km2)) |>
  mutate(mireInSurvery_percent = round(mireInSurvey_km2/mire_km2*100, 2))

@@

@@

To describe these two concepts we use five condition variables from the same data set ( @tbl-variables).
The data set is from a standardised field survey of nature types in Norway that started in 2018 and which is still ongoing.
The data is freely available [@norwegian_environmental_agency_naturtyper_2024].
See @miljødirektoratet for the sampling protocol (in Norwegian).
We included data from 2018 to 2023.
In this survey, selected nature types are delineated on a map, and each locality is scored on a range of variables relevant for describing the state and quality of nature.
The surveys are commissioned with the goal of producing data relevant for immediate land-use decisions, and is therefore spatially biased, typically towards areas with high human impact or expected impact.
In addition there is a thematic and size bias in the sampling protocol.
Of all possible mire types, the survey only maps the following:

@@

```{r}
#| eval: true

muni_tbl <- readRDS("../data/municipality_stats.RDS")

@@

| Municipality | Total terrestrial area (km^2^)                                            | \% of terrestrial area surveyed                                            | \% open mires in relation to total terrestrial area                         | Total mire area (km^2^)                                                  | %of mire area inside survey area                                                      | Number of mire polygons in survey                                 | Mean Infrastructure Index value  |
|-------|----------|----------|----------|----------|----------|----------|----------|
| Nordre Follo | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(t_area_km)` | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(dk_percent)` | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(mirePercent)` | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(mire_km2)` | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(mireInSurvery_percent)` | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(n)` |                                  |
| Trondheim    | `r muni_tbl |> filter(Municipality == "Trondheim") |> pull(t_area_km)`    | `r muni_tbl |> filter(Municipality == "Trondheim") |> pull(dk_percent)`    |                                                                             |                                                                          |                                                                                       | `r muni_tbl |> filter(Municipality == "Trondheim") |> pull(n)`    |                                  |
| Målselv      | `r muni_tbl |> filter(Municipality == "Målselv") |> pull(t_area_km)`      | `r muni_tbl |> filter(Municipality == "Målselv") |> pull(dk_percent)`      |                                                                             |                                                                          |                                                                                       | `r muni_tbl |> filter(Municipality == "Målselv") |> pull(n)`      |                                  |

>>>>

---
title: Using spatially biased variables in ecosystem condition accounting with a GIS based workflow
subtitle: A Short Subtitle
author:
  - name: Anders Lorentzen Kolstad
    email: anders.kolstad@nina.no
    affiliations: 
        - id: some-tech
          name: Norwegian Institute for Nature Research
          department: Department of Terrestrial Ecology
          address: Street Address
          city: City
          state: State
          postal-code: Postal Code
    attributes:
        corresponding: true
    note: This is the first author footnote.
  - name: Bob Security
    email: bob@example.com
    affiliations:
        - id: another-u
          name: Another University
          department: Department Name
          address: Street Address
          city: City
          state: State
          postal-code: Postal Code
    note: |
      Another author footnote, this is a very long footnote and it should be a really long footnote. But this footnote is not yet sufficiently long enough to make two lines of footnote text.
  - name: Cat Memes
    email: cat@example.com
    affiliations:
        - ref: another-u
    note: Yet another author footnote.
  - name: Derek Zoolander
    email: derek@example.com
    affilations:
        - ref: some-tech
abstract: |
  This is the abstract. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Vestibulum augue turpis, dictum non malesuada a, volutpat eget velit. Nam placerat turpis purus, eu tristique ex tincidunt et. Mauris sed augue eget turpis ultrices tincidunt. Sed et mi in leo porta egestas. Aliquam non laoreet velit. Nunc quis ex vitae eros aliquet auctor nec ac libero. Duis laoreet sapien eu mi luctus, in bibendum leo molestie. Sed hendrerit diam diam, ac dapibus nisl volutpat vitae. Aliquam bibendum varius libero, eu efficitur justo rutrum at. Sed at tempus elit.
keywords: 
  - SEEA EA
  - alien species
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: Ecological Indicators
      formatting: preprint
      model: 3p
      cite-style: authoryear
    include-in-header:
      text: '\usepackage{lineno}\linenumbers'
editor: 
  markdown: 
    wrap: sentence
# set eval to true to actually run the analysis
execute:
  echo: false
  include: false
  eval: false
  warning: false
  message: false
---

<!--# The document is set to eval=F in the yaml, and some code chunks are set manually to eval=T. When working on the manuscript, you can run all (Ctr+Alt+R). This will read inn data cashed locally. To also evaluate the more time consuming chunks, change the YAML to eval: true and run all (or render to html):  -->

```{r setup}
#| warning: false
#| message: false
#| include: false
#| eval: true

library(tidyverse)
library(knitr)
library(sf)
library(tmap)
library(stars)
library(terra)
library(tidyterra)
library(ggtext)

myCRS <- 25832
```

```{r paths}
#| eval: true

# Conditional file directory
dir <- substr(getwd(), 1,2)

# Soem directories
mireFile <- "P:/41201785_okologisk_tilstand_2022_2023/data/Myrmodell/myrmodell90pros.tif"

path_infrastructure <- ifelse(dir == "C:", 
        "R:/GeoSpatialData/Utility_governmentalServices/Norway_Infrastructure_Index/Original/Infrastrukturindeks_UTM33/infra_tiff.tif",
        "/data/R/GeoSpatialData/Utility_governmentalServices/Norway_Infrastructure_Index/Original/Infrastrukturindeks_UTM33/infra_tiff.tif")

path_naturetypes <- ifelse(dir == "C:", 
               "R:/GeoSpatialData/Habitats_biotopes/Norway_Miljodirektoratet_Naturtyper_nin/Original/Naturtyper_nin_0000_norge_25833_FILEGDB/Naturtyper_nin_0000_norge_25833_FILEGDB.gdb",
               "/data/R/GeoSpatialData/Habitats_biotopes/Norway_Miljodirektoratet_Naturtyper_nin/Original/Naturtyper_nin_0000_norge_25833_FILEGDB/Naturtyper_nin_0000_norge_25833_FILEGDB.gdb")

path_muni <- paste0("../data/Basisdata_0000_Norge_25833_Kommuner_FGDB.gdb")

@@

@@

# The data is too big to be stored on GitHub

# Impart polygon data set
naturetypes <- sf::st_read(dsn = path_naturetypes)
# 117k polygons (2023)
# sometimes this functin struggles with Norwegian letters. 
# Restart R and try again.

@@

# 13k obs.

naturetypes |>
  saveRDS("../data/surveyData.RDS")
#head(naturetypes)
```

@@

```{r getMunicipalities}
#| eval: true

# Preparing the outlines for the three municipalieties

# find the correct layer
#st_layers(path_muni)

# read inn data and transform
muni <- sf::read_sf(path_muni, layer = "kommune") |>
  st_transform(myCRS)

# This returns some multisurfaces as well
#table(st_geometry_type(muni))
# function to make sure that multipolygons are returned 
ensure_multipolygons <- function(X) {
    tmp1 <- tempfile(fileext = ".gpkg")
    tmp2 <- tempfile(fileext = ".gpkg")
    st_write(X, tmp1)
    gdalUtilities::ogr2ogr(tmp1, tmp2, f = "GPKG", nlt = "MULTIPOLYGON")
    Y <- st_read(tmp2)
    st_sf(st_drop_geometry(X), geom = st_geometry(Y))
}

@@

# To crop EDM, I need the three municipalities seprately. 
nf <- muni3 |>
  filter(kommunenummer == "3020")
tr <- muni3 |>
  filter(kommunenummer == "5001")
me <- muni3 |>
  filter(kommunenummer == "5418")
```

# I need to intersect the naturetypes data with the municipalities
nature3 <- naturetypes |>
  st_intersection(muni3)

@@

# and also to get the data coverage polygon. 
# These I had to manually download from naturbase.no 
dk <-   read_sf("../data/dekningskartNF/Naturtyper NiN - Dekningskart.shp") |>
  rbind(read_sf("../data/dekningskartTR/Naturtyper NiN - Dekningskart.shp")) |>
  rbind(read_sf("../data/dekningskartME/Naturtyper NiN - Dekningskart.shp")) |> 
  st_transform(st_crs(muni3)) |>
  st_intersection(muni3) |>
  st_union()

```{r prepSomeMoreMunicipalityShapes}
#| eval: true

@@

# calculate stats - terrestrial area
terrestrial <- terrestrial |>
  mutate(area_t = geometry |> st_area(),
         t_area_km = 
           round(units::drop_units(area_t * 1e-6)))|>
  as_tibble()
```

```{r positionMap}

# Make map to show where the three municipalities are
# First, get centroids
centroids <- muni3 |>
  st_centroid()

tmap_mode("plot")
positionMap <- tm_shape(outline)+
  tm_polygons() +
  tm_shape(centroids)+
    tm_text(text = "Municipality",
      size = 1,
      xmod = 2,
      ymod=-1)+
  tm_symbols(size = 2,
             shape = 17,
             col="black")+
  tm_grid(projection = 4326)+
  tm_layout(bg.color = "skyblue")
```

@@

# Testing the cropping proceedure using a downsampled starts object 
mire_test <- read_stars(mireFile, proxy = T) |>
  setNames("mire") |>
  st_downsample(100) |>
  mutate(mire = na_if(mire, 0)) |>
  st_as_stars()

@@

# The cropping is correct:
#tmap_mode("plot")
#tm_shape(mire_test)+
#  tm_raster(palette = "red")+
#  tm_shape(outline)+
#  tm_borders()+
#  tm_shape(mire_test_nf)+
#  tm_raster(col="red")

# Then we do the same, but without downsampling

mire_nf <- read_stars(mireFile, proxy = T) |>
  setNames("mire") |>
  #mutate(mire = na_if(mire, 0)) |>
  st_crop(nf) |>
  st_as_stars()
# The cropped raster is for some reason far into the ocean!

# The cells have values, so it seems like there is something 
# that happens after the cropping
tmap_mode("plot")
tm_shape(mire_test)+
  tm_raster(palette = "red")+
  tm_shape(outline)+
  tm_borders()+
  tm_shape(mire_nf)+
  tm_raster(col="red")
```

```{r mireTerra}

# trying { terra } instead

# convert municipal outline to vect via st
nf_vect <- as(nf, "Spatial") |>
  terra::vect()

@@

```{r dk2-CoverageMaps}
# calculate area of survey coverage maps
# values goes into summary table in the ms
dk2 <-   read_sf("../data/dekningskartNF/Naturtyper NiN - Dekningskart.shp") |>
  rbind(read_sf("../data/dekningskartTR/Naturtyper NiN - Dekningskart.shp")) |>
  rbind(read_sf("../data/dekningskartME/Naturtyper NiN - Dekningskart.shp")) |>
  st_transform(st_crs(muni3)) |>
  st_intersection(muni3) |>
  st_cast("POLYGON") |>
  group_by(Municipality) |>
  summarise(geometry = st_union(geometry)) |>
  mutate(dk_area_km = geometry |> st_area(),
         dk_area_km = round(units::drop_units(dk_area_km * 1e-6)))
```

```{r mireArea}
# calculate area of the mires in each municipality
# -- Nordre Follo
mireArea <- mire_nf_terra |>
  global(c("mean", "sum"), na.rm=T) |>
  add_column("Municipality" = "Nordre Follo") |>
  mutate(mirePercent = round(mean*100, 1),
         mire_km2 = sum/1e+4)

# Caululate the area of mire inside the coverage maps
# -- Nordre Follo
mire_nf_terra_dk <- mire_nf_terra |>
  terra::mask(dk2 |> filter(Municipality == "Nordre Follo")) |>
  global("sum", na.rm=T) |>
  mutate(mireInSurvey_km2 = sum/1e+4) |>
  add_column(Municipality = "Nordre Follo")

@@

```{r infrastructureIndex}
# this data is on a 100x100m grid
infra <- stars::read_stars(path_infrastructure) |>
  # , which is mor then we need - warp it to 1x1km
  st_warp(cellsize = c(1000, 1000), 
                                   crs = st_crs(nf), 
                                   use_gdal = TRUE,
                                   method = "average") |>
  setNames("infrastructureIndex") |>
  st_transform(myCRS) |>
   mutate(infrastructureIndex = case_when(
    infrastructureIndex < 1 ~ 0,
    infrastructureIndex < 6 ~ 1,
    infrastructureIndex < 12 ~ 2,
    infrastructureIndex >= 12 ~ 3)) |>
  # takeing away point in the sea
  st_crop(outline)

@@

# path to local cahing folder
path_temp <- ifelse(dir == "C:", 
        "P:/41201785_okologisk_tilstand_2022_2023/data/cache/",
        "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/")

saveRDS(infra, paste0(path_temp, "infrastructureIndex_discrete_vectorized.rds"))
```

```{r readInfraStructureIndex}

# read cached vetorized infrastructure data
infra <- readRDS(paste0(path, "infrastructureIndex_discrete_vectorized.rds"))

# intersect with the three municipalities
# and calculate area
infraMuni3 <- infra |>
  st_intersection(muni3) |>
  mutate(area = geometry |> st_area())

# Turn m2 into km2
# and sum the total area per HIA
(infraMuni3_tbl <- infraMuni3 |>
  as.data.frame() |>
  mutate(area_HIA_km2 = units::drop_units(area)*1e-6) |>
  group_by(Municipality, infrastructureIndex) |>
  summarise(total_area_HIAs_km2 = round(sum(area_HIA_km2))))

# Calculate the area weighted mean HIA value per municipality
infraMuni3_summary <- infraMuni3_tbl |>
  group_by(Municipality) |>
  summarise(meanHIA = 
              round(
                weighted.mean(
                  infrastructureIndex, total_area_HIAs_km2), 2))

# Make a plot to check that it has worked 
infraMuni3_tbl |>
  ggplot()+
  geom_bar(aes(x = infrastructureIndex, 
               y = total_area_HIAs_km2,
               fill = factor(infrastructureIndex),
               colour = factor(infrastructureIndex)),
           stat = "identity",
           lwd=1.2
           )+
  scale_fill_manual(values = RColorBrewer::brewer.pal(4, "YlOrBr"))+
  scale_color_manual(values = RColorBrewer::brewer.pal(5, "YlOrBr")[-1])+
  facet_grid(cols = vars(Municipality))+
  theme_minimal_hgrid()+
  labs(x = "Human Impact Area (HIA) class",
       y = "Area (km<sup>2</sup>)")+
  theme(axis.title.x = element_textbox_simple(
      width = NULL,
      padding = margin(4, 4, 4, 4),
      margin = margin(4, 0, 0, 0),
      linetype = 1,
      r = grid::unit(8, "pt"),
      fill = "azure1"
    ),
    axis.title.y = element_textbox_simple(
      width = NULL,
      padding = margin(4, 4, 4, 4),
      margin = margin(4, 0, 0, 0),
      linetype = 1,
      orientation = "left-rotated",
      r = grid::unit(8, "pt"),
      fill = "azure1"),
    strip.background = element_blank(),
    strip.text = element_textbox(
      size = 12,
      color = "white", fill = "#5D729D", box.color = "#4A618C",
      halign = 0.5, linetype = 1, r = unit(5, "pt"), width = unit(1, "npc"),
      padding = margin(2, 0, 1, 0), margin = margin(3, 3, 3, 3)
    )
    )+
guides(fill= "none", colour = "none")
```

<!--# Need to update this figure -->

```{r infraMuniMaps}
infraMuniMap <- tm_shape(muni3)+
  tm_borders()+
tm_shape(infraMuni3)+
  tm_polygons(col= "infrastructureIndex",
              style="cat",
              title = "Human Impact\nArea Class")+
  tm_shape(muni3)+
  tm_borders(lwd=3, col="black")+
  tm_facets(by = "Municipality")+
  tm_layout(legend.show=TRUE)

# Plot
muniPlot <- tm_shape(muni3)+
    tm_borders()+
  tm_facets(by = "Municipality",
            ncol=3)+
  tm_shape(terrestrial)+
    tm_fill(col = "lightgreen",
             alpha=.4)+
  tm_shape(ocean)+
    tm_polygons(col = "skyblue",
                border.col = "black")+
  tm_shape(dk)+
    tm_polygons(col = "grey",
                alpha=.8)+
  tm_shape(nature3)+
  tm_polygons(col = "red",
              border.col = "red",
              lwd=3)

(methodsMap <- tmap_arrange(
                          #positionMap, 
                           muniPlot,
                          infraMuniMap,
             widths = c(.4, .6)))

#cowplot::plot_grid(tmap_grob(muniPlot),
#                   tmap_grob(infraMuniMap),
#                   ncol=1,)

```{r muni-tbl}

# Calulate some more stats for the table with municipality stats

# Number of mire polygons per municipality
# -- Try count() here instead
nature3_tbl <- nature3 |>
  group_by(Municipality) |>
  summarise(n = n()) |>
  as_tibble()

# Take muni3 and add all sorts of other data to it
# using left_join.
# Save local cached file
muni_tbl <- muni3 |>
  # calculate total area
  mutate(area_km = 
           round(
             units::drop_units(
               geom |> st_area() * 1e-6))) |>
  # make tibble for the coming join
  as_tibble() |>
  # paste inn terrestrial area
  left_join(terrestrial |> select(kommunenummer, t_area_km),
            keep=F) |>
  # area of survey
  left_join(dk2 |> select(Municipality, dk_area_km)) |>
  mutate(dk_percent = round((dk_area_km/t_area_km) * 100)) |>
  # number of polygons
  left_join(nature3_tbl |> select(Municipality, n)) |>
  # total mire area and %
  left_join(mireArea |> select(Municipality, mirePercent, mire_km2)) |>
  # % mire inside survey coverage map
  left_join(mire_nf_terra_dk |> select(Municipality, mireInSurvey_km2)) |>
  mutate(mireInSurvery_percent = round(mireInSurvey_km2/mire_km2*100, 2)) |>
  left_join(infraMuni3_summary)

@@

```{r read-munitbl}
#| eval: true
muni_tbl <- readRDS("../data/municipality_stats.RDS")
```

@@

To describe these two characteristics we use five condition variables from the same data set ( @tbl-variables).
The data set is from a standardised field survey of nature types in Norway that started in 2018 and which is still ongoing.
The data is freely available [@norwegian_environmental_agency_naturtyper_2024].
See @miljødirektoratet for the sampling protocol (in Norwegian).
We included data from 2018 to 2023.
In this survey, selected nature types are delineated on a map, and each locality is scored on a range of variables relevant for describing the state and quality of nature.
The surveys are commissioned with the goal of producing data relevant for immediate land-use decisions, and is therefore spatially biased, typically towards areas with high human impact or expected impact.
In addition there is a thematic and size bias in the sampling protocol.
Of all possible mire types, the survey only maps the following:

@@

@@

| Municipality | Total terrestrial area (km^2^)                                            | \% of terrestrial area surveyed                                            | \% open mires in relation to total terrestrial area                         | Total mire area (km^2^)                                                  | %of mire area inside survey area                                                      | Number of mire polygons in survey                                 | Mean Infrastructure Index value                                         |
|---------|---------|---------|---------|---------|---------|---------|---------|
| Nordre Follo | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(t_area_km)` | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(dk_percent)` | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(mirePercent)` | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(mire_km2)` | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(mireInSurvery_percent)` | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(n)` | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(meanHIA)` |
| Trondheim    | `r muni_tbl |> filter(Municipality == "Trondheim") |> pull(t_area_km)`    | `r muni_tbl |> filter(Municipality == "Trondheim") |> pull(dk_percent)`    |                                                                             |                                                                          |                                                                                       | `r muni_tbl |> filter(Municipality == "Trondheim") |> pull(n)`    | `r muni_tbl |> filter(Municipality == "Trondheim") |> pull(meanHIA)`    |
| Målselv      | `r muni_tbl |> filter(Municipality == "Målselv") |> pull(t_area_km)`      | `r muni_tbl |> filter(Municipality == "Målselv") |> pull(dk_percent)`      |                                                                             |                                                                          |                                                                                       | `r muni_tbl |> filter(Municipality == "Målselv") |> pull(n)`      | `r muni_tbl |> filter(Municipality == "Målselv") |> pull(meanHIA)`      |
